Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-05-25 hazelnut-popl17.tex
wrote figures for dynamics
progress on intro

@@

\maketitle
\begin{abstract}
Well-typed expressions are rich inductive structures, but programmers typically construct and manipulate them only indirectly, as text that must pass through a parser and typechecker. Not all text survives this journey. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. Contending with (i.e. reasoning about) malformed text or well-formed but meaningless expressions is a difficult task, for programmers and their tools alike.

@@

We introduce Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms. Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems. %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

>>>>

\usepackage{todonotes}

@@

\maketitle
\begin{abstract}
Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker. Not all text survives this journey. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. 
% Not all text that arises during the programming process survives the journey through a parser and typechecker. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed. 
Contending with (i.e. reasoning about) malformed text, or well-formed but meaningless expressions, is a difficult task, for programmers and tools alike.

@@

This paper introduces Hazelnut, a minimal \emph{structure editor} where every edit action leaves the expression being edited in both a statically and dynamically meaningful state (with the latter implying that Hazelnut is particularly suitable as a basis for \emph{live programming}.) Uniquely, 1) Hazelnut is defined in a principled type-theoretic style, with a rich metatheory mechanized in Agda; and 2) users need not construct outer forms before they construct inner forms. Instead, Hazelnut's \emph{action semantics} automatically places terms that have a type that is inconsistent with the expected type into a \emph{hole}, which safely defers the type consistency check until the hole is \emph{finished}. Interestingly, this mechanism is related to a mechanism that arises in the semantics of gradual type systems. %Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

### 2016-05-25 intro.tex
progress on intro

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use are often presented with text that does not correspond to a well-defined program. This may be because the programmer is in the midst of a sequence of
text edit actions that leave the program ``temporarily'' malformed or ill-defined, or because the programmer has made a mistake. This complicates the programming process, both because the language definition provides no reasoning principles relevant to such text, and, relatedly, because the editor can no longer provide useful services (e.g. syntax highlighting, or semantics-aware code completion.) %Tool designers often develop \emph{ad hoc} heuristic methods that allow one to

Source code editors sometimes develop \emph{ad hoc} workarounds for this problem, e.g. they might use whitespace to guess where a construct is likely to end, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.)

>>>>

%Although all programs can be represented as text, most text does not correspond to a syntactically well-formed and semantically well-defined program. 
Programmers and the tools that they use often encounter text that does not correspond to a meaningful program. This may be because the programmer is in the midst of a sequence of 
edit actions that leaves the text temporarily malformed or ill-typed, or because the programmer has made a mistake. It is difficult for tools to help programmers execute a corrective course of action in such situations (e.g. by providing syntax highlighting and semantics-aware code completion.) %because the language definition is necessarily silent about the structure or meaning of such text. %For example, a code editor cannot assist the programmer in these situations.

Some editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might use whitespace to guess where a construct is likely to end, attempt to insert closing delimiters automatically, or recover from a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.\todo{citations?}) These heuristic methods are quite complex and sometimes misleading.
